---
title: "NDVI Green Up Date Calculator"
author: "Ryan McKeon"
date: "1/17/2020"
output: html_document
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# IMPORT Libraries
library(data.table)
library(lubridate)
library(changepoint)
library(zoo)
library(gridExtra)
library(tidyverse)


```
## What this project does

This code answers a question I was curious about last summer... "Is spring starting earlier than it used to?"  Given ample evidence for a warming climate, I figured I already knew the answer, but it was the challenge of actually acquiring, processing, and visualizing data to inform my hunch that I was most interested in.

#### The Data -- Normalized Difference Vegetation Index 
This project works with data gathered through **Google Earth Engine** (GEE) using JS script available here and a user-specified analysis region which can be created in GEE. Specifically it uses spectral imagery from the MODIS Terra satellite (8-Day Global 500m product) to calculate a time series of the Normalized Difference Vegetation Index (**NDVI**). NDVI values range from -1 to 1 and indicate the density of photosynthesizing vegetation on the surface through variations in the adsorption and reflectance of near infrared and red bands of solar energy throughout the year. This example is for the Upper Valley region of New Hampshire and Vermont.

#### The Processing -- How to define the start of spring?
We begin with a time series of NDVI starting in 2000 when the satellite became operational and extending to the first week of 2020.  This code walks through processing this time series to identify the timing of the start of the spring “green up” (when grass starts growing and leaves pop out of deciduous trees and bushes) for each year of data.  Along the way it discusses how to contend with noise and uncertainty in the data and the rationale for decisions about how the data was processed and interpreted.  

 
## Getting Started
Loading data and initial processing and visualization. 

```{r, echo=FALSE}
# load MODIS NDVI data
ndviDF = fread("UV_poly_ndvi_2000_2020.csv") 

# convert "Date" to date object AND put it in a new field called "rDate"
ndviDF$rDate = parse_date_time(ndviDF$Satellite_data, "%d-%m-%y")

# remove rows NAs from the data set
ndviDF  = ndviDF[is.na(ndviDF$ndviMedian) == FALSE, ]

# add year field by reading the date-time in "rDate" field
ndviDF$year = year(ndviDF$rDate)

# Plot the NDVI Time Series
ggplot(data=ndviDF) +
  geom_line(aes(x=ndviDF$rDate, y=ndviDF$ndviMedian)) +
  xlab("Year") +
  ylab("Median NDVI Observed") +
  ggtitle("NDVI from MODIS Terra 8-day Global 500m Data Set")

```

**The NDVI time series for the Upper Valley Region**<br>
From this plot one can easily see the annual cycle of vegetation on this New England landscape and some of the curiosities of working with NDVI data.  First, the summers all max out at a fairly stable maximum “greeness” value, but the winters are not nearly as consistent, which is likely due to varying amounts of snow on the ground.  The relative sameness of the summers vs. the winters is easy to see on the histogram of the same data below...

```{r, echo=F}
# Histogram of all NDVI values observed
ggplot(data=ndviDF) +
  geom_histogram(aes(ndviMedian), binwidth = 0.03) +
  #geom_vline(xintercept = 0.3, color = "blue") +
  #geom_vline(xintercept = 0.8, color = "red") +
  xlab("Median NDVI Observed") +
  ylab("Count") +
  ggtitle("Distribution of All NDVI Values Observed")
```

**20 Year Histogram of NDVI for the Upper Valley Region**<br>

Here the winter values spread out over a much wider range at the low end of the NDVI spectrum than the summer values that build a very high peak at the top of range.  For the task at hand, we are going to ignore the high end of the specturm almost entirely, because it is the rollover from winter to spring that I need to find to answer my question up above.


```{r small_multiples, echo=F}
# Work with Data at Annual Scale... Inflection Point analysis  --- First Visualize

# remove 2020 because it is incomplete
ndviDF = ndviDF[ndviDF$year<2020]

# Make a Facet Figure (Small Multiples) showing median NDVI for each year and Loess Smoothing
ggplot(data = ndviDF, aes(x=pixelJdateMode, y=ndviMedian)) +
  geom_point(aes(y=ndviMedian), color="black", size=1) +
  #geom_hline(yintercept = 0.235, color = "blue", size=0.5) +          # draw a horizonal line on the plot
  #geom_hline(yintercept = 0.6, color = "green", size=0.5) +
  facet_wrap(~year) +
  xlab("Julian Date") +
  ylab("Median NDVI") +
  ggtitle("NDVI Data By Year") +
  theme_light()






```


Looking at all of the years of data...



```{r loess_smoothing, echo=F}

# Great Resource for Smoothing methods... https://rafalab.github.io/dsbook/smoothing.html

# How to set the SPAN parameter of loess smoothing (this controls the proportion of the number of data points that contribute to the fit at a given point)
# consider the data... 1 point in a 8 day period
# consider the process... what is the time scale of major changes in NDVI -- seasonal?  3 months per season in 12 months per year...

seasonSpan = 3/12  # n months of thing of interest / total months in data set

# subset the input data to keep only rows from a single year and filter out NAs in ndvi
ndviYear <- ndviDF %>%
  filter(year == 2018)

# use Loess() to smooth data -- Here using a 50% Span factor
#loess50 = loess(ndviMedian ~ pixelJdateMode, data = ndviYear, span = .5)
#ndviYear$loess50 = predict(loess50)


# plot up the single year of data using the Julian Date of pixel data acquisition 
ggplot(data = ndviYear, aes(x=pixelJdateMode, y=ndviMedian)) +
  geom_point(color = "black", size = 2) +
  geom_smooth(method="loess", span=1, method.args = list(degree=1), mapping=aes(color="dark gray"), se=F) +   # loess smoothing Linear
  geom_smooth(method="loess", span=seasonSpan+seasonSpan, method.args = list(degree=1), mapping=aes(color="light blue"), se=F) +   # loess smoothing Linear fitting
  geom_smooth(method="loess", span=.125, method.args = list(degree=1), mapping=aes(color="pink"), se=F) +   # loess smoothing Linear
  geom_smooth(method="loess", span=seasonSpan, method.args = list(degree=1), mapping=aes(color="red"), se=F) +   # loess smoothing Linear fitting
  xlab("Julian Date") +
  ylab("NDVI") +
  ggtitle("NDVI for 2018 with Loess Smoothing") +
  scale_color_manual(                #  Manually Build a Legend for the colors in the aes() above
    name= "Span Setting", 
    values= c("red"="red","pink"="pink","light blue"="light blue","dark gray"="dark gray"),
    labels= c("12 months","6 months","1.5 months", "3 months")) +
  theme_light()



```

```{r}
# Work with Data at Annual Scale... FIND Greatest Green UP acceleration
# First figure it out for a single year...  Later put it in a loop for all years

Year = 2002

# subset the input data to keep only rows from a single year and filter out NAs in ndvi
ndviYear <- ndviDF %>%
  filter(year == Year & is.na(ndviMedian) == FALSE)

# use Loess() to smooth data -- Here using a 25% Span factor based on the length of seasons -- see above
loessSeason = loess(ndviMedian ~ pixelJdateMode, data = ndviYear, span = seasonSpan, degree = 1) # use a linear fit instead of a polynomial fit
ndviYear$loess = predict(loessSeason)

# Find slope of loess smoothed fit to time series
ndviYear$slope = 0
for (i in 2:nrow(ndviYear)) {
  ndviYear[i,"slope"] = (ndviYear[i,"loess"] - ndviYear[i-1,"loess"])/(ndviYear[i,"pixelJdateMode"] - ndviYear[i-1,"pixelJdateMode"])
}

# find the highest positive slope value (i.e. greening up in spring) and store the Julian Date
maxSlope = max(ndviYear$slope)
maxSlopeJdate = ndviYear[ndviYear$slope == maxSlope,]$pixelJdateMode  

# Calc rate of change of slope (acceleration of greening) of loess smoothed fit
ndviYear$acceleration = 0
for (i in 3:nrow(ndviYear)) {
  ndviYear[i,"acceleration"] = (ndviYear[i,"slope"] - ndviYear[i-1,"slope"])/(ndviYear[i,"pixelJdateMode"] - ndviYear[i-1,"pixelJdateMode"])
}

# Find the highest acceleration prior to the highest slope

# Use slope and acceleration of greening to estimate the start of "Green Up"
# Here the "Start of Green Up" is defined as the highest acceleration prior to the steepest slope AND slope is positive, i.e. it is greening up
AccelerationDF <- ndviYear %>%
  filter(pixelJdateMode < maxSlopeJdate & slope > 0)

maxSpringAcceleration = max(AccelerationDF$acceleration)
maxSpringAccJdate = ndviYear[ndviYear$acceleration == maxSpringAcceleration,]$pixelJdateMode  


# plot up the single year of data using the Julian Date of pixel data acquisition 
p1 = ggplot(data = ndviYear) +
  geom_point(mapping = aes(x=pixelJdateMode, y=ndviMedian),color = "black",, shape = 21, size = 1.5) +
  geom_smooth(mapping = aes(x=pixelJdateMode, y=ndviMedian), method="loess", span=seasonSpan, method.args = list(degree=1), color="red", se=TRUE) +
  geom_vline(xintercept = maxSpringAccJdate, color = "red") +
  geom_vline(xintercept = maxSlopeJdate, color = "blue") +
  xlab("") +
  ylab("Smoothed NDVI") +
  theme_light() 

p2 = ggplot(data = ndviYear) +
 
  geom_col(mapping = aes(x=pixelJdateMode, y=slope*100), color = "dark blue", size = 3, alpha = 0.6) +
  geom_vline(xintercept = maxSpringAccJdate, color = "red") +
  geom_vline(xintercept = maxSlopeJdate, color = "blue") +
  xlab("") +
  ylab("Slope") +
  theme_light()

p3 = ggplot(data = ndviYear) +
  geom_col(mapping = aes(x=pixelJdateMode, y=acceleration*1000), color = "dark red", size = 1.5, alpha = 0.8) +
  geom_vline(xintercept = maxSpringAccJdate, color = "red") +
  geom_vline(xintercept = maxSlopeJdate, color = "blue") +
  xlab("Julian Date") +
  ylab("Acceleration") +
  theme_light()

grid.arrange(p1,p2,p3, ncol = 1)

```
## Find the green up date for both methods for the whole time series

```{r}

# Set up loop that calculates the date of the steepest slope and greatest acceleration for each year in the record

seasonSpan = 3/12  # n months of thing of interest / total months in data set
greenUpDF = setNames(data.frame(matrix(ncol = 7, nrow = 20)), c("year", "greenUpJdateSlope", "greenUpJdateAcc", "ndviMedianSlope", "ndviMedianAcc", "maxSlope", "maxAcceleration"))

for (i in 2000:2019) {
  currentYear = i
  currentIDX = i-1999
  
  # subset the input data to keep only rows from a single year and filter out NAs in ndvi
  ndviYear <- ndviDF %>%
    filter(year == currentYear & is.na(ndviMedian) == FALSE)
  
# use Loess() to smooth data -- Here using a 25% Span factor based on the length of seasons -- see above
loessSeason = loess(ndviMedian ~ pixelJdateMode, data = ndviYear, span = seasonSpan, degree = 1) # use a linear fit instead of a polynomial fit
ndviYear$loess = predict(loessSeason)

# Find slope of loess smoothed fit to time series
ndviYear$slope = 0
for (i in 2:nrow(ndviYear)) {
  ndviYear[i,"slope"] = (ndviYear[i,"loess"] - ndviYear[i-1,"loess"])/(ndviYear[i,"pixelJdateMode"] - ndviYear[i-1,"pixelJdateMode"])
}

# find the highest positive slope value (i.e. greening up in spring) and store the Julian Date
maxSlope = max(ndviYear$slope)
maxSlopeJdate = ndviYear[ndviYear$slope == maxSlope,]$pixelJdateMode  

# Calc rate of change of slope (acceleration of greening) of loess smoothed fit
ndviYear$acceleration = 0
for (i in 3:nrow(ndviYear)) {
  ndviYear[i,"acceleration"] = (ndviYear[i,"slope"] - ndviYear[i-1,"slope"])/(ndviYear[i,"pixelJdateMode"] - ndviYear[i-1,"pixelJdateMode"])
}

# Find the highest acceleration prior to the highest slope

# Use slope and acceleration of greening to estimate the start of "Green Up"
# Here the "Start of Green Up" is defined as the highest acceleration prior to the steepest slope AND slope is positive, i.e. it is greening up
AccelerationDF <- ndviYear %>%
  filter(pixelJdateMode < maxSlopeJdate & slope > 0)

maxSpringAcceleration = max(AccelerationDF$acceleration)
maxSpringAccJdate = ndviYear[ndviYear$acceleration == maxSpringAcceleration,]$pixelJdateMode  
  
 
  greenUpDF[currentIDX,"year"] = currentYear
  greenUpDF[currentIDX,"greenUpJdateSlope"] = maxSlopeJdate
  greenUpDF[currentIDX,"greenUpJdateAcc"] = maxSpringAccJdate
  greenUpDF[currentIDX,"ndviMedianSlope"] = ndviYear[ndviYear$slope == maxSlope,]$ndviMedian
  greenUpDF[currentIDX,"ndviMedianAcc"] = ndviYear[ndviYear$acceleration == maxSpringAcceleration,]$ndviMedian
  greenUpDF[currentIDX,"maxSlope"] = maxSlope
  greenUpDF[currentIDX,"maxAcceleration"] = maxSpringAcceleration
}
  
ggplot(greenUpDF) +
  geom_point(mapping = aes(x=year, y=greenUpJdateSlope), shape= 21, size=3, fill = "blue") +
  geom_point(mapping = aes(x=year, y=greenUpJdateAcc), shape=21, size=3,  fill = "red") +
  geom_smooth(mapping = aes(x=year, y=greenUpJdateSlope), method="lm", color="blue") +
  geom_smooth(mapping = aes(x=year, y=greenUpJdateAcc), method="lm", color="red") +
  xlab("Year") +
  ylab("Green Up Date (Julian)") +
  ggtitle("The Start of Spring Through the Years")
  




```

